\center \fontsize{18.8}{20} \selectfont Towards Globally Color-Dual All-Loop Integrands

### 2023-09-19 notes/2loopNLSM.tex
added citations - my hands are off

  We catalog and analyze ?4? theories which all manifest the duality
  between color and kinematics both at tree and one-loop level.  We
  identify a unifying theory, \emph{semi-abelian} Yang--Mills, from
  which all four of the theories can be extracted.  Joint
  consideration of these theories highlights that the lifting of
  one-loop color-kinematics dual numerators to two loops may be just
  as challenging as the step from tree level to one loop.  As a first
  step in understanding the two-loop challenge, we construct
  color-dual two-loop four-point non-linear sigma model (NLSM)
  numerators using generalized unitarity methods.  The double-copy of
  the constructed NLSM numerators with themselves, $\mathcal{N}=0$
  Yang--Mills, and $\mathcal{N}=4$ Yang--Mills correctly reproduce the
  known physical unitarity cuts of special Gallileons, Born-Infeld
  theory, and Dirac-Born-Infeld-Volkov-Akulov theory, respectively.
  These NLSM numerators are in explicit tension with generalizations
  of the one-loop NLSM numerators. \ace{I realize that we all know
    this, but does/will a statement to this effect actually make it
    into the draft?}  Additionally, we revisit the two-loop four-point
  Yang--Mills problem in order to identify the exact point of
  inconsistency in a local color-dual construction.

@@

\setstackgap{S}{6pt}
\setstackgap{L}{7pt}
\newpage
\section{Introduction}\label{intro}
Since the turn of the century, our understanding of the $S$-matrix and its hidden structures has expanded dramatically. At the heart of this progress is the mantra that physical observables are simpler when studied on-shell. However, while plugging in on-shell states certainly offers a path towards taming the factorial growth of Feynman diagrams, it also obscures the off-shell simplicity at the heart of many quantum field theories. A shining example of this hidden structure is the duality between color and kinematics \cite{BCJ}, which states that the kinematic numerators of gauge theory amplitudes can be rearranged to obey the same algebraic identities as the constituent color factors. When this duality is realized globally off-shell, it dramatically decreases the combinatorial complexity posed by integrand construction, and offers a path towards efficient assembly of quantum gravity integrands directly from simpler gauge theory building blocks via the double copy \cite{}.

However, despite the tremendous success in leveraging this duality to compute gauge and gravity observables to high orders in perturbation theory \cite{}, color-kinematics remains a conjecture at loop level. Indeed, there are many examples in the literature where identifying color-dual representations beyond one-loop has posed a formidable challenge \cite{}. In this work, we use the nonlinear sigma model (NLSM) and Yang-Mills, two theories proven to permit color-dual representations at tree-level \cite{}, as case studies for advancing our understanding of the kinematic algebra to the multi-loop level.

@@

Faced with this obstruction, we study the multi-loop sector of both NLSM and Yang-Mills in \cref{2loopBoot} using an ansatz based color-dual bootstrap. In contrast to much of the available literature on construction of color-dual integrands at two-loop \cite{}, our results are completely agnostic to the spacetime dimension; all the polarizations and momenta appearing in our construction will remain formally $D$-dimensional. This makes our methods and results particularly well suited for algorithmically extracting rational terms from the $D$-dependence of dimensionally regulated loop momenta. We are able to achieve this by employing a custom solver, \texttt{FiniteFieldSolve}, which will soon be made public.

With this solver, we are able to compute a two-loop integrand for NLSM that globally manifests the duality off-shell. This represents a new benchmark for color-dual representations in non-supersymmetric gauge theories. However in contrast, we find surprisingly that Yang-Mills does not permit a globally color-dual representation, even when considering the most general polynomial ansatz of Lorentz covariant kinematics. Concretely, we find a sharp failure point between the ``bowtie'' cut, and the following jacobi triple:
\begin{eBox}
\begin{equation}
   \NMCutC
  \Rightarrow
  \JacobiOneDoubleBox +  \JacobiOneCrossedBox + \JacobiOnePentaTriangle \neq 0 \,.
\end{equation}
\end{eBox}
We comment on the implications of this finding in \cref{conclusions}, and discuss alternative approaches where one must relax field theoretic locality in order to accommodate the construction of globally color-dual multi-loop integrands.

@@

Many of the 1-loop results presented in \cref{sec:cubic} are informed by the tree-level double-copy so we will review the tree case here.
The color structure of tree-level amplitudes is particularly simple.
A fully color dressed tree amplitude $\mathcal{A}_n$ in YM, NLSM, or bi-adjoint scalar theory can be decomposed in terms of the trace basis
\be
\mathcal{A}_n = \sum \limits_{\sigma\in S_{n-1}} \text{Tr}(T^{a_{\sigma_1}} T^{a_{\sigma_2}}...T^{a_{\sigma_{n-1}}} T^{a_n}) A_n[\sigma_1, \sigma_2,...\sigma_{n-1}, n],
\ee
where $T^a$ are generators of the gauge group and, because of the cyclicity of the trace, one of the legs is held fixed so that the sum runs over the $(n-1)!$ permutations of the remaining external legs.
The coefficients of the color factors are the color-ordered partial amplitudes $A_n[...]$.
Any tree-level color factor can be converted into a linear combination of Del Duca-Dixon-Maltoni (DDM) half-ladder color factors
\be
\text{DDM}[a_1, a_2,... a_n] \equiv f^{a_1 a_2 b_1} f^{b_1 a_3 b_2} f^{b_2 a_4 b_3}...f^{b_{n-3} a_{n-1} a_n}
\ee
associated with the graph
\be
\halfLadder
\ee
by repeated application of the Jacobi identity \cite{DixonMaltoni}.
The fully color-dressed tree amplitude can then be re-expressed as
\be
\mathcal{A}_n = \sum \limits_{\sigma\in S_{n-2}} \text{DDM}[a_1, a_{\sigma_2}, a_{\sigma_3}, a_{\sigma_{n-1}}, a_n]  A_n[1, \sigma_2,...\sigma_{n-1}, n]
\ee
with legs $1$ and $n$ fixed so that the sum only contains $(n-2)!$ terms, indicating that there are additional relations -- the Kleiss-Kuijf (KK) relations -- amongst the partial amplitudes \cite{Kleiss:1988ne}.
The KK relations are generic to any theory with purely adjoint particles.
Color-kinematics duality implies that the KK basis is overcomplete and can be reduced to a basis of $(n-3)!$ amplitudes via the fundamental BCJ identities \cite{BCJ}
\be
\sum \limits_{i=2}^{n-1} k_1 \cdot (k_2+...+k_i) A_n[2,...,i,1,i+1,...,n] =0.
\ee
In terms of constructing explicit color-dual solutions at tree level, it is enough to specify the numerator of the half ladder because any diagram can be reduced to a half ladder through successive applications of the Jacobi identity.
At 1-loop, every diagram can be related to one topology as well, in this case the $n$-gon, but the pattern does not persist to 2-loops.

\subsection{Color-dual Lagrangians}\label{offShellCK}
While BCJ relations is an on-shell statement about the color-dual nature of scattering amplitudes, one might aspire to hard code color-kinematics duality by realizing an off-shell Lagrangian description of the theory. The double-copy construction of \ref{DCconstruction} would suggest a manifestly cubic Lagrangian of the form:
\be
\mathcal{L} = P_V^{ab}P_W^{ij}\mathcal{O}^a_i \Box \mathcal{O}^b_j +  V^{abc} W^{ijk}\mathcal{O}^a_i \mathcal{O}^b_j \mathcal{O}^c_k
\ee
where $V^{abc}$ and $W^{ijk}$ are cubic Feynman rules mixing field operators $\mathcal{O}^{a}_i$ indexed by quantum numbers $a$ and $i$. We have introduced $P_V^{ab}$ and $P_W^{ij}$ projection operators to encode non-local kinetic structure that could in principle participate in the construction. Should the off-shell Feynman rules of this cubic theory be antisymmetric and satisfy Jacobi identities on all internal edges,
\begin{align}
\textit{anti-symmetry}:& \quad{}_{a_1}\langle V^{a_2} \rangle_{a_3}+\text{cyc}(23)=0\,,
\\
\textit{Jacobi identity}:& \quad {}_{a_1}\langle V^{a_2} V^{a_3}\rangle_{a_4}+\text{cyc}(234)=0\,,
\end{align}
where 
\be
{}_{a_1}\langle V^{a_2}V^{a_3}\cdots V^{a_{n-1}} \rangle_{a_n} \equiv V^{a_1a_2b_1}V^{b_1a_3b_2}\cdots V^{b_{n-3} a_{n-2}a_n}\,.
\ee
This is a sufficient condition of the amplitudes to obey BCJ relations. Off-shell description of color-dual theories are rare, and only a few examples are available in the literature. Typically when one goes about construction color-dual theories directly from a set of cubic interactions, higher point Jacobi and antisymmetry constraints require introducing additional operators. See refs \cite{} for some recent examples. Indeed, as we'll show in \cref{sec:cubic}, Yang-Mills requires introducing a cubic 2-form interaction in order to manifest the kinematic algebra just at four-point. Absent information about the kinematic algebra, one can make progress in the construction of loop-level color-dual amplitudes by employing a color-dual bootstrap. 
\subsection{Color-dual Bootstrap}
\label{sec:bootstrap}

@@

We can achieve this construction by splitting the gauge group of the full Yang-Mills action from $G \rightarrow G_L \times G_R$, where the full non-abelian gauge symmetry is restored only when we identify $G_L = G_R$. However, if we break the left gauge group down to $G_L \rightarrow U(1)^{N^2}$, and take the right gauge group to $G_R \rightarrow U(N)$, then we are left with a semi-abelian theory that is gauge invariant in the preffered $U(1)$-directions. We call this manifestly color-dual theory, \textit{semi-abelian Yang-Mills},
\begin{eBox}
\be
\label{eq:semiYM}
\mathcal{L}^{\text{semi-YM}} = -\frac{1}{2}\text{tr}\left[\bar{F}_{\mu\nu}F^{\mu\nu}\right]
\ee
\end{eBox}
where,
\be
\bar{F}^a_{\mu\nu} = \partial_\mu \bar{A}_\nu^a - \partial_\nu \bar{A}_\mu^a
\ee
\be
F^a_{\mu\nu} = \partial_\mu A_\nu^a - \partial_\nu A_\mu^a + f^{abc} A_\mu^b A_\nu^c .
\ee
Here we can think of the abelian vector as a background field that sources on-shell $A_\mu$ currents. Indeed, as we describe in the Appendix \ref{}, semi-abelian Yang-Mills theory is at the heart of YZ-theory \cite{}, J-theory \cite{}, and self-dual Yang-Mills \cite{}. The Feynman rules fo this theory are simply:
\be
\cubic{hgrey0}{fermion2}{}{fermion2}{}{fermion2}{}= (\varepsilon_1\bar{\varepsilon}_3)(\varepsilon_2 k_3) -  (\varepsilon_2\bar{\varepsilon}_3)(\varepsilon_1 k_3)
\ee
Specifically, semi-Abelian YM is simply a clever reinterpretation of the Lagrangian obtained by integrating an auxiliary field into the J-theory equations of motion.
We can use the Ward identity of \ref{eq:GIrelA} to check that the amplitudes are indeed gauge invariant with the abelian polarizations are taken on-shell, $\bar{\varepsilon} \rightarrow k$.

@@

Our first application of this theory for color-dual construction at loop level is for self-dual Yang-Mills (SDYM). At tree-level, the off-shell cubic vertices can be written in terms of light cone coordinates \cite{}:

@@

where $\mathcal{O}_i$ are the basis tensors of \cite{}, and $ \mathcal{R}_i$ is a rational function of momenta of the form;

>>>>

  We make progress towards understanding how color-dual representations must be constructed at the multi-loop level. First
  identify a cubic theory, \emph{semi-abelian} Yang--Mills, that unifies many of the color-dual theories constructed in the literature, and provides a construction of color-dual numerators at one-loop directly from Feynman rules. Joint
  consideration of these theories highlights that the lifting of
  one-loop color-kinematics dual numerators to two loops may be just
  as challenging as the step from tree level to one loop.  As a first
  step in understanding the two-loop challenge, we construct
  color-dual two-loop four-point non-linear sigma model (NLSM)
  numerators using generalized unitarity methods.  The double-copy of
  the constructed NLSM numerators with themselves, $\mathcal{N}=0$
  Yang--Mills, and $\mathcal{N}=4$ Yang--Mills correctly reproduce the
  known physical unitarity cuts of special Gallileons, Born-Infeld
  theory, and Dirac-Born-Infeld-Volkov-Akulov theory, respectively.
  These NLSM numerators are in explicit tension with generalizations
  of the one-loop NLSM numerators. \ace{I realize that we all know
    this, but does/will a statement to this effect actually make it
    into the draft?}  Additionally, we revisit the two-loop four-point
  Yang--Mills problem in order to identify the exact point of
  inconsistency in a local color-dual construction.

@@

\setstackgap{S}{6pt}
\setstackgap{L}{7pt}

\section{Introduction}\label{intro}
Since the turn of the century, our understanding of the $S$-matrix and its hidden structures has expanded dramatically. At the heart of this progress is the mantra that physical observables are simpler when studied on-shell \cite{Parke:1986gb,TasiLance,Cheung:2017pzi}. However, while plugging in on-shell states certainly offers a path towards taming the factorial growth of Feynman diagrams, it also obscures the off-shell simplicity at the heart of many quantum field theories. A shining example of this hidden structure is the duality between color and kinematics \cite{BCJ,Bern:2010ue,BCJreview}, which states that the kinematic numerators of gauge theory amplitudes can be rearranged to obey the same algebraic identities as the constituent color factors. When this duality is realized globally off-shell, it dramatically decreases the combinatorial complexity posed by integrand construction, and offers a path towards efficient assembly of quantum gravity integrands directly from simpler gauge theory building blocks via the double copy \cite{BCJ,Bern:2010ue}.

However, despite the tremendous success in leveraging this duality to compute gauge and gravity observables to high orders in perturbation theory \cite{FiveLoopN4,GeneralizedDoubleCopyFiveLoops,Bern:2018jmv,Carrasco:2021otn}, color-kinematics remains a conjecture at loop level. Indeed, there are many examples in the literature where identifying color-dual representations beyond one-loop has posed a formidable challenge \cite{Mogull:2015adi,Johansson:2017bfl,KalinN2TwoLoop}. In this work, we use the nonlinear sigma model (NLSM) and Yang-Mills, two theories proven to permit color-dual representations at tree-level \cite{Feng:2010my,Cachazo:2014xea}, as case studies for advancing our understanding of the kinematic algebra to the multi-loop level.

@@

Faced with this obstruction, we study the multi-loop sector of both NLSM and Yang-Mills in \cref{2loopBoot} using an ansatz based color-dual bootstrap. In contrast to much of the available literature on construction of color-dual integrands at two-loop \cite{Bern:2013yya,Mogull:2015adi,Johansson:2017bfl}, our results are completely agnostic to the spacetime dimension; all the polarizations and momenta appearing in our construction will remain formally $D$-dimensional. This makes our methods and results particularly well suited for algorithmically extracting rational terms from the $D$-dependence of dimensionally regulated loop momenta. We are able to achieve this by employing a custom solver, \texttt{FiniteFieldSolve}, which will soon be made public.

With this solver, we are able to compute a two-loop integrand for NLSM that globally manifests the duality off-shell. This represents a new benchmark for color-dual representations in non-supersymmetric gauge theories. However in contrast, we find surprisingly that Yang-Mills does not permit a globally color-dual representation, even when considering the most general polynomial ansatz of Lorentz covariant kinematics. Concretely, we find a sharp failure point between the ``bowtie'' cut, and the following jacobi triple:
\begin{equation}
   \NMCutC
  \Rightarrow
  \JacobiOneDoubleBox +  \JacobiOneCrossedBox + \JacobiOnePentaTriangle \neq 0 \,.
\end{equation}
We comment on the implications of this finding in \cref{conclusions}, and discuss alternative approaches where one must relax field theoretic locality in order to accommodate the construction of globally color-dual multi-loop integrands.

@@

Many of the 1-loop results presented in \cref{sec:cubic} are informed by the tree-level double-copy so we will review the tree case here.
The color structure of tree-level amplitudes is particularly simple.
A fully color dressed tree amplitude $\mathcal{A}_n$ in YM, NLSM, or bi-adjoint scalar theory can be decomposed in terms of the trace basis
\be
\mathcal{A}_n = \sum \limits_{\sigma\in S_{n-1}} \text{Tr}(T^{a_{\sigma_1}} T^{a_{\sigma_2}}...T^{a_{\sigma_{n-1}}} T^{a_n}) A_n[\sigma_1, \sigma_2,...\sigma_{n-1}, n],
\ee
where $T^a$ are generators of the gauge group and, because of the cyclicity of the trace, one of the legs is held fixed so that the sum runs over the $(n-1)!$ permutations of the remaining external legs.
The coefficients of the color factors are the color-ordered partial amplitudes $A_n[...]$.
Any tree-level color factor can be converted into a linear combination of Del Duca-Dixon-Maltoni (DDM) half-ladder color factors
\be
\text{DDM}[a_1, a_2,... a_n] \equiv f^{a_1 a_2 b_1} f^{b_1 a_3 b_2} f^{b_2 a_4 b_3}...f^{b_{n-3} a_{n-1} a_n}
\ee
associated with the graph
\be
\halfLadder
\ee
by repeated application of the Jacobi identity \cite{DixonMaltoni}.
The fully color-dressed tree amplitude can then be re-expressed as
\be
\mathcal{A}_n = \sum \limits_{\sigma\in S_{n-2}} \text{DDM}[a_1, a_{\sigma_2}, a_{\sigma_3}, a_{\sigma_{n-1}}, a_n]  A_n[1, \sigma_2,...\sigma_{n-1}, n]
\ee
with legs $1$ and $n$ fixed so that the sum only contains $(n-2)!$ terms, indicating that there are additional relations -- the Kleiss-Kuijf (KK) relations -- amongst the partial amplitudes \cite{Kleiss:1988ne}.
The KK relations are generic to any theory with purely adjoint particles.
Color-kinematics duality implies that the KK basis is overcomplete and can be reduced to a basis of $(n-3)!$ amplitudes via the fundamental BCJ identities \cite{BCJ,Feng:2010my}
\be
\sum \limits_{i=2}^{n-1} k_1 \cdot (k_2+...+k_i) A_n[2,...,i,1,i+1,...,n] =0.
\ee
In terms of constructing explicit color-dual solutions at tree level, it is enough to specify the numerator of the half ladder because any diagram can be reduced to a half ladder through successive applications of the Jacobi identity.
At 1-loop, every diagram can be related to one topology as well, in this case the $n$-gon, but the pattern does not persist to 2-loops.

\subsection{Color-dual Lagrangians}\label{offShellCK}
While BCJ relations is an on-shell statement about the color-dual nature of scattering amplitudes, one might aspire to hard code color-kinematics duality by realizing an off-shell Lagrangian description of the theory. The double-copy construction of \ref{DCconstruction} would suggest a manifestly cubic Lagrangian of the form:
\be
\mathcal{L} = P_V^{ab}P_W^{ij}\mathcal{O}^a_i \Box \mathcal{O}^b_j +  V^{abc} W^{ijk}\mathcal{O}^a_i \mathcal{O}^b_j \mathcal{O}^c_k
\ee
where $V^{abc}$ and $W^{ijk}$ are cubic Feynman rules mixing field operators $\mathcal{O}^{a}_i$ indexed by quantum numbers $a$ and $i$. We have introduced $P_V^{ab}$ and $P_W^{ij}$ projection operators to encode non-local kinetic structure that could in principle participate in the construction. Should the off-shell Feynman rules of this cubic theory be antisymmetric and satisfy Jacobi identities on all internal edges,
\begin{align}
\textit{anti-symmetry}:& \quad{}_{a_1}\langle V^{a_2} \rangle_{a_3}+\text{cyc}(23)=0\,,
\\
\textit{Jacobi identity}:& \quad {}_{a_1}\langle V^{a_2} V^{a_3}\rangle_{a_4}+\text{cyc}(234)=0\,,
\end{align}
where 
\be
{}_{a_1}\langle V^{a_2}V^{a_3}\cdots V^{a_{n-1}} \rangle_{a_n} \equiv V^{a_1a_2b_1}V^{b_1a_3b_2}\cdots V^{b_{n-3} a_{n-2}a_n}\,.
\ee
This is a sufficient condition of the amplitudes to obey BCJ relations. Off-shell description of color-dual theories are rare, and only a few examples are available in the literature. Typically when one goes about construction color-dual theories directly from a set of cubic interactions, higher point Jacobi and antisymmetry constraints require introducing additional operators. See refs \cite{Ben-Shahar:2022ixa} for some recent examples. Indeed, as we'll show in \cref{sec:cubic}, Yang-Mills requires introducing a cubic 2-form interaction in order to manifest the kinematic algebra just at four-point. Absent information about the kinematic algebra, one can make progress in the construction of loop-level color-dual amplitudes by employing a color-dual bootstrap. 
\subsection{Color-dual Bootstrap}
\label{sec:bootstrap}

@@

We can achieve this construction by splitting the gauge group of the full Yang-Mills action from $G \rightarrow G_L \times G_R$, where the full non-abelian gauge symmetry is restored only when we identify $G_L = G_R$. However, if we break the left gauge group down to $G_L \rightarrow U(1)^{N^2}$, and take the right gauge group to $G_R \rightarrow U(N)$, then we are left with a semi-abelian theory that is gauge invariant in the preffered $U(1)$-directions. We call this manifestly color-dual theory, \textit{semi-abelian Yang-Mills},
\begin{eBox}
\be
\label{eq:semiYM}
\mathcal{L}^{\text{semi-YM}} = -\frac{1}{2}\text{tr}\left[\bar{F}_{\mu\nu}F^{\mu\nu}\right]
\ee
\end{eBox}
where,
\be
\bar{F}^a_{\mu\nu} = \partial_\mu \bar{A}_\nu^a - \partial_\nu \bar{A}_\mu^a
\ee
\be
F^a_{\mu\nu} = \partial_\mu A_\nu^a - \partial_\nu A_\mu^a + f^{abc} A_\mu^b A_\nu^c .
\ee
Here we can think of the abelian vector as a background field that sources on-shell $A_\mu$ currents. Indeed, as we describe in the Appendix \cref{sec:CKLagrangians}, semi-abelian Yang-Mills theory is at the heart of YZ-theory \cite{Cheung:2016prv}, J-theory \cite{Cheung:2020djz,Cheung:2021zvb}, and self-dual Yang-Mills \cite{Monteiro2011pc}. The Feynman rules fo this theory are simply:
\be
\cubic{hgrey0}{fermion2}{}{fermion2}{}{fermion2}{}= (\varepsilon_1\bar{\varepsilon}_3)(\varepsilon_2 k_3) -  (\varepsilon_2\bar{\varepsilon}_3)(\varepsilon_1 k_3)
\ee
Specifically, semi-Abelian YM is simply a clever reinterpretation of the Lagrangian obtained by integrating an auxiliary field into the J-theory equations of motion.
We can use the Ward identity of \ref{eq:GIrelA} to check that the amplitudes are indeed gauge invariant with the abelian polarizations are taken on-shell, $\bar{\varepsilon} \rightarrow k$.

@@

Our first application of this theory for color-dual construction at loop level is for self-dual Yang-Mills (SDYM). At tree-level, the off-shell cubic vertices can be written in terms of light cone coordinates \cite{Monteiro2011pc}:

@@

where $\mathcal{O}_i$ are the basis tensors of \cite{Bern:2017tuc,Carrasco:2019yyn}, and $ \mathcal{R}_i$ is a rational function of momenta of the form;
